---
title: "R Notebook"
output: 
  bookdown::html_document2:
    toc: true
author: "David Brown & Emily DeJeu"
bibliography: mda_bib.bib
link-citations: yes
nocite: |
  @*
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, setup, include=FALSE}
knitr::opts_knit$set(
  root.dir = rprojroot::find_rstudio_root_file(),
  collapse = TRUE,
  comment = "#>",
  tidy.opts=list(width.cutoff=70),
  tidy=TRUE
)
```

# Introduction

The code 

## Multi-Dimensional Analysis (MDA)

This project uses an approach called Multi-Dimensional Analysis (MDA). MDA is a [complex statistical procedure](https://www.uni-bamberg.de/fileadmin/eng-ling/fs/Chapter_21/Index.html?Multidimensionalanalysis.html) developed by Douglas Biber. It is largely used to describe language as it varies by [genre, register, and use](https://www.uni-bamberg.de/fileadmin/eng-ling/fs/Chapter_21/Index.html?21Summary.html).

MDA is based on the fundamental linguistic principle that some linguistic variables co-occur (nouns and adjectives, for example) while others inversely co-occur (think nouns and pronouns).

MDA in conducted in the following stages:

1. Identification of relevant variables
1. Extraction of factors from variables
1. Functional interpretation of factors as dimensions
1. Placement of categories on the dimensions

## Corpora

Our study compares disciplinary variation in student writing across two national contexts: American and British. The American data come from the Michigan Corpus of Upper-Level Student Papers (MICUSP), while the British come from the British Academic Written English (BAWE) corpus.

# Comparing MICUSP & BAWE

## Load Packages

MDA is carried out using the **mda.biber** packager. You can access the package and its code here:

<https://github.com/browndw/mda.biber>

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(mda.biber)
```

## Load Utility Functions

```{r}
source("R/utils.R")
```

## Load & Prepare Data

### Load Tables of DocuScope Frequencies

```{r message=FALSE}
ds_micusp <- read_csv("data/CLUSTER_micusp_body.csv", show_col_types = FALSE)
micusp_meta <- read_csv("data/micusp_meta.csv", show_col_types = FALSE)
```

```{r}
ds_micusp[1:10, 1:6] %>% knitr::kable(digits = 3, caption = 'Sample of DocuScope data tables.') %>%  kableExtra::kable_styling()
```

### Load Metadata

```{r message=FALSE}
ds_bawe <- read_csv("data/CLUSTER_bawe_body.csv", show_col_types = FALSE)
bawe_meta <- read_csv("data/bawe_meta.csv", show_col_types = FALSE)
```

## Composition of Corpora

```{r warning=FALSE, message=FALSE}
micusp_counts <- ds_micusp %>%
  mutate(Filename = str_remove(Filename, ".txt$")) %>%
  rename(text_id = Filename) %>%
  left_join(select(micusp_meta, text_id, paper_discipline)) %>%
  group_by(paper_discipline) %>%
  summarize(Tokens = sum(Tokens),
    Texts = n()
  )

```

```{r echo=FALSE}
 micusp_counts %>% 
  janitor::adorn_totals() %>%
  knitr::kable("html",
               col.names = c("", "Tokens", "Texts"),
               format.args = list(big.mark = ",", scientific = FALSE),
               caption = 'Composition of MICUSP.') %>% 
  kableExtra::row_spec(18, bold = TRUE) %>%
  kableExtra::kable_styling()
```


```{r warning=FALSE, message=FALSE}
bawe_counts <- ds_bawe %>%
  mutate(Filename = str_remove(Filename, ".txt$")) %>%
  rename(text_id = Filename) %>%
  left_join(select(bawe_meta, text_id, discipline)) %>%
  group_by(discipline) %>%
  summarize(Tokens = sum(Tokens),
            Texts = n()
  )
```

```{r echo=FALSE}
 bawe_counts %>% 
  janitor::adorn_totals() %>% 
  knitr::kable("html",
               col.names = c("", "Tokens", "Texts"),
               format.args = list(big.mark = ",", scientific = FALSE),
               caption = 'Composition of BAWE') %>% 
  kableExtra::row_spec(32, bold = TRUE) %>%
  kableExtra::kable_styling()
```

## Prepare Data

```{r message=FALSE}
ds_micusp <- ds_micusp %>%
  mutate(Filename = str_remove(Filename, ".txt")) %>%
  rename(text_id = Filename) %>%
  left_join(select(micusp_meta, text_id, paper_discipline)) %>%
  column_to_rownames("text_id") %>%
  select(paper_discipline, AcademicTerms:Updates) %>%
  mutate(paper_discipline = as.factor(paper_discipline))

ds_bawe <- ds_bawe %>%
  mutate(Filename = str_remove(Filename, ".txt")) %>%
  rename(text_id = Filename) %>%
  left_join(select(bawe_meta, text_id, discipline)) %>%
  column_to_rownames("text_id") %>%
  select(discipline, AcademicTerms:Updates) %>%
  mutate(discipline = as.factor(discipline))
```

## Execute MDA

### MICUSP MDA

```{r}
mda_micusp <- mda_loadings(ds_micusp, n_factors = 3)

mda_bawe <- mda_loadings(ds_bawe, n_factors = 3)
```

### Examine MICUSP Loadings

```{r}
micusp_fl <- attributes(mda_micusp)$loadings %>% 
  rownames_to_column("Category") %>%
  pivot_longer(!Category, names_to = "Factor", values_to = "MICUSP Loading")

bawe_fl <- attributes(mda_bawe)$loadings %>% 
  rownames_to_column("Category") %>%
  pivot_longer(!Category, names_to = "Factor", values_to = "BAWE Loading")

fl_1 <- full_join(micusp_fl, bawe_fl, by = c("Category", "Factor")) %>%
  filter(abs(`MICUSP Loading`) > 0.35 | abs(`BAWE Loading`) > 0.35) %>%
  mutate(Category = str_replace_all(Category, "([a-z])([A-Z])", "\\1 \\2")) %>%
  mutate(Factor = str_replace (Factor, "(\\d)", " \\1")) %>%
  select(Factor, Category, `MICUSP Loading`) %>%
  arrange(Factor, -`MICUSP Loading`)

fl_2 <- full_join(micusp_fl, bawe_fl, by = c("Category", "Factor")) %>%
  filter(abs(`MICUSP Loading`) > 0.35 | abs(`BAWE Loading`) > 0.35) %>%
  mutate(Category = str_replace_all(Category, "([a-z])([A-Z])", "\\1 \\2")) %>%
  mutate(Factor = str_replace (Factor, "(\\d)", " \\1")) %>%
  select(Factor, Category, `BAWE Loading`) %>%
  arrange(Factor, -`BAWE Loading`)

micusp_orp <- which(abs(fl_1$`MICUSP Loading`) < 0.35 & fl_1$`MICUSP Loading` > 0)
bawe_orp <- which(abs(fl_2$`BAWE Loading`) < 0.35 & fl_2$`BAWE Loading` > 0)

micusp_orn <- which(abs(fl_1$`MICUSP Loading`) < 0.35 & fl_1$`MICUSP Loading` < 0)
bawe_orn <- which(abs(fl_2$`BAWE Loading`) < 0.35 & fl_2$`BAWE Loading` < 0)

micusp_neg <- which(fl_1$`MICUSP Loading` < 0 & abs(fl_1$`MICUSP Loading`) > 0.35)
bawe_neg <- which(fl_2$`BAWE Loading` < 0 & abs(fl_2$`BAWE Loading`) > 0.35)
```

```{r echo=FALSE, warning=FALSE}
fl_1 %>%
  knitr::kable(digits = 2, caption = 'Factor loadings for MICUSP.') %>%
  kableExtra::kable_styling(full_width = F, position = "float_left") %>%
  kableExtra::pack_rows(index = table(fl_1$Factor)) %>%
  kableExtra::row_spec(micusp_neg, bold = T, color = "white", background = "steelblue") %>%
  kableExtra::row_spec(micusp_orp, background = "#d3d3d3") %>%
  kableExtra::row_spec(micusp_orn, bold = T, color = "white", background = "#809bbf") %>%
  kableExtra::remove_column(1)

 fl_2 %>%
  knitr::kable(digits = 2, caption = 'Factor loadings for BAWE.') %>%
  kableExtra::kable_styling(full_width = F, position = "right") %>%
  kableExtra::pack_rows(index = table(fl_2$Factor)) %>%
  kableExtra::row_spec(bawe_neg, bold = T, color = "white", background = "steelblue") %>%
  kableExtra::row_spec(bawe_orp, background = "#d3d3d3") %>%
  kableExtra::row_spec(bawe_orn, bold = T, color = "white", background = "#809bbf") %>%
  kableExtra::remove_column(1)

```

### Vizualizing Dimensional Composition Using Radar Plots

```{r}
radar_1 <- plot_radar(mda_micusp, mda_bawe, factor_n = 1)
```

```{r fig.height=5, fig.width=7.5, echo=FALSE, fig.cap="Compositions of Dimension 1 for BAWE and MICUSP."}
radar_1
```

```{r evaluate=FALSE, echo=FALSE}
ggsave("output/radar_1.png", radar_1,  height = 5, width = 5, units = "in", dpi = 300)
```


```{r}
radar_2 <- plot_radar(mda_micusp, mda_bawe, factor_n = 2)
```

```{r fig.height=5, fig.width=7.5, echo=FALSE, fig.cap="Compositions of Dimension 2 for BAWE and MICUSP."}
radar_2
```

```{r evaluate=FALSE, echo=FALSE}
ggsave("output/radar_2.png", radar_1,  height = 5, width = 5, units = "in", dpi = 300)
```

```{r}
radar_3 <- plot_radar(mda_micusp, mda_bawe, factor_n = 3)
```

```{r fig.height=5, fig.width=7.5, echo=FALSE, fig.cap="Compositions of Dimension 3 for BAWE and MICUSP."}
radar_3
```

```{r evaluate=FALSE, echo=FALSE}
ggsave("output/radar_3.png", radar_1,  height = 5, width = 5, units = "in", dpi = 300)
```


### Evaluate Using ANOVA

```{r}
# Carry out regression
fit1 <- lm(Factor1 ~ group, data = mda_micusp)
fit2 <- lm(Factor1 ~ group, data = mda_bawe)

# Convert ANOVA results into data.frames allows for easier name manipulation
f1_aov <- data.frame(anova(fit1), r.squared = c(summary(fit1)$r.squared*100, NA))
f2_aov <- data.frame(anova(fit2), r.squared = c(summary(fit2)$r.squared*100, NA))

# Putting all into one data.frame/table
anova_results <- data.frame(rbind(c("DF", "Sum Sq", "Mean Sq", "F value", "Pr(>F)", "*R*^2^", 
                                    "DF", "Sum Sq", "Mean Sq", "F value", "Pr(>F)", "*R*^2^"), 
                                   cbind(round(f1_aov, 2), round(f2_aov, 2)))) 
colnames(anova_results) <- c("", "", "", "", "", "", "", "", "", "", "", "")
row.names(anova_results)[1] <- ""
anova_results[is.na(anova_results)] <- "--"
```

```{r}
anova_results %>% knitr::kable("html",
               caption = 'Dimension 1 ANOVAs.') %>% 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = T) %>%
  kableExtra::add_header_above(c("", "MICUSP (Dim 1)" = 6, "BAWE (Dim 1)" = 6))
```


```{r echo=FALSE}
fit1 <- lm(Factor2 ~ group, data = mda_micusp)
fit2 <- lm(Factor2 ~ group, data = mda_bawe)

f1_aov <- data.frame(anova(fit1), r.squared = c(summary(fit1)$r.squared*100, NA))
f2_aov <- data.frame(anova(fit2), r.squared = c(summary(fit2)$r.squared*100, NA))

anova_results <- data.frame(rbind(c("DF", "Sum Sq", "Mean Sq", "F value", "Pr(>F)", "*R*^2^", 
                                    "DF", "Sum Sq", "Mean Sq", "F value", "Pr(>F)", "*R*^2^"), 
                                   cbind(round(f1_aov, 2), round(f2_aov, 2)))) 
colnames(anova_results) <- c("", "", "", "", "", "", "", "", "", "", "", "")
row.names(anova_results)[1] <- ""
anova_results[is.na(anova_results)] <- "--"

anova_results %>% knitr::kable("html",
               caption = 'Dimension 2 ANOVAs.') %>% 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = T) %>%
  kableExtra::add_header_above(c("", "MICUSP (Dim 2)" = 6, "BAWE (Dim 2)" = 6))
```


```{r echo=FALSE}
fit1 <- lm(Factor3 ~ group, data = mda_micusp)
fit2 <- lm(Factor3 ~ group, data = mda_bawe)

f1_aov <- data.frame(anova(fit1), r.squared = c(summary(fit1)$r.squared*100, NA))
f2_aov <- data.frame(anova(fit2), r.squared = c(summary(fit2)$r.squared*100, NA))

anova_results <- data.frame(rbind(c("DF", "Sum Sq", "Mean Sq", "F value", "Pr(>F)", "*R*^2^", 
                                    "DF", "Sum Sq", "Mean Sq", "F value", "Pr(>F)", "*R*^2^"), 
                                   cbind(round(f1_aov, 2), round(f2_aov, 2)))) 
colnames(anova_results) <- c("", "", "", "", "", "", "", "", "", "", "", "")
row.names(anova_results)[1] <- ""
anova_results[is.na(anova_results)] <- "--"

anova_results %>% knitr::kable("html",
               caption = 'Dimension 3 ANOVAs.') %>% 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = T) %>%
  kableExtra::add_header_above(c("", "MICUSP (Dim 3)" = 6, "BAWE (Dim 3)" = 6))
```

### Examine Position of Disciplines along Dimensions

```{r}
p1 <- stickplot_dual(mda_micusp, mda_bawe, title_a = "MICUSP", title_b = "BAWE")
```

```{r fig.height=5, fig.width=7.5, echo=FALSE, fig.cap="Position of disciplines along Dimension 1."}
p1
```

```{r eevaluate=FALSE, echo=FALSE}
ggsave("output/dim_1.png", p1,  height = 5, width = 7.5, units = "in", dpi = 300)
```


```{r}
p2 <- stickplot_dual(mda_micusp, mda_bawe, n_factor = 2, title_a = "MICUSP", title_b = "BAWE")
```

```{r fig.height=5, fig.width=7.5, echo=FALSE}
p2
```

```{r evaluate=FALSE, echo=FALSE}
ggsave("output/dim_2.png", p2,  height = 5, width = 7.5, units = "in", dpi = 300)
```

```{r}
p3 <- stickplot_dual(mda_micusp, mda_bawe, n_factor = 3, title_a = "MICUSP", title_b = "BAWE")
```

```{r fig.height=5, fig.width=7.5, echo=FALSE}
p3
```

```{r evaluate=FALSE, echo=FALSE}
ggsave("output/dim_3.png", p2,  height = 5, width = 7.5, units = "in", dpi = 300)
```

# Compare DocuScope and Biber

## Load & Prepare Data

```{r}
biber_micusp <- read_csv("data/BIBER_micusp_body.csv", show_col_types = FALSE)
```

```{r}
biber_micusp <- biber_micusp %>%
  mutate(doc_id = str_remove(doc_id, ".txt")) %>%
  rename(text_id = doc_id) %>%
  left_join(select(micusp_meta, text_id, paper_discipline)) %>%
  column_to_rownames("text_id") %>%
  select(paper_discipline, f_01_past_tense:f_67_neg_analytic) %>%
  mutate(paper_discipline = as.factor(paper_discipline))
```

## Execute MDA

```{r}
mda_biber <- mda_loadings(biber_micusp, n_factors = 3)
```

### Re-label Factors

```{r}
colnames(mda_biber) <- c("group", "Factor2", "Factor1", "Factor3")
colnames(attributes(mda_biber)$group_means) <- c("group", "Factor2", "Factor1", "Factor3")
colnames(attributes(mda_biber)$loadings) <- c("Factor2", "Factor1", "Factor3")
```

## Examine Factor Loadings

```{r}
micusp_fl <- attributes(mda_micusp)$loadings %>% 
  rownames_to_column("Category") %>%
  pivot_longer(!Category, names_to = "Factor", values_to = "MICUSP Loading") %>%
  filter(abs(`MICUSP Loading`) > 0.35) %>%
  mutate(Category = str_replace_all(Category, "([a-z])([A-Z])", "\\1 \\2")) %>%
  mutate(Factor = str_replace (Factor, "(\\d)", " \\1")) %>%
  arrange(Factor, -`MICUSP Loading`)

biber_fl <- attributes(mda_biber)$loadings %>% 
  rownames_to_column("Category") %>%
  pivot_longer(!Category, names_to = "Factor", values_to = "BIBER Loading") %>%
  filter(abs(`BIBER Loading`) > 0.35) %>%
  mutate(Category = str_remove(Category, "f_\\d+_")) %>%
  mutate(Category = str_replace_all(Category, "_", " ")) %>%
  mutate(Category = str_to_title(Category)) %>%
  mutate(Factor = str_replace (Factor, "(\\d)", " \\1")) %>%
  arrange(Factor, -`BIBER Loading`)

micusp_neg <- which(micusp_fl$`MICUSP Loading` < 0 & abs(micusp_fl$`MICUSP Loading`) > 0.35)
biber_neg <- which(biber_fl$`BIBER Loading` < 0 & abs(biber_fl$`BIBER Loading`) > 0.35)

```


```{r echo=FALSE, warning=FALSE}
micusp_fl %>%
  knitr::kable(digits = 2, caption = 'Factor loadings for DocuScope') %>%
  kableExtra::kable_styling(full_width = F, position = "float_left") %>%
  kableExtra::pack_rows(index = table(micusp_fl$Factor))%>%
  kableExtra::row_spec(micusp_neg, bold = T, color = "white", background = "steelblue")

 biber_fl %>%
  knitr::kable(digits = 2, caption = 'Factor loadings for Biber') %>%
  kableExtra::kable_styling(full_width = F, position = "right") %>%
  kableExtra::pack_rows(index = table(biber_fl$Factor)) %>%
  kableExtra::row_spec(biber_neg, bold = T, color = "white", background = "steelblue")

```

## Evaluate Using ANOVA

```{r}
# Carry out regression
fit1 <- lm(Factor1 ~ group, data = mda_micusp)
fit2 <- lm(Factor1 ~ group, data = mda_biber)

# Convert ANOVA results into data.frames allows for easier name manipulation
f1_aov <- data.frame(anova(fit1), r.squared = c(summary(fit1)$r.squared*100, NA))
f2_aov <- data.frame(anova(fit2), r.squared = c(summary(fit2)$r.squared*100, NA))

# Putting all into one data.frame/table
anova_results <- data.frame(rbind(c("DF", "Sum Sq", "Mean Sq", "F value", "Pr(>F)", "*R*^2^", 
                                    "DF", "Sum Sq", "Mean Sq", "F value", "Pr(>F)", "*R*^2^"), 
                                  cbind(round(f1_aov, 2), round(f2_aov, 2)))) 
colnames(anova_results) <- c("", "", "", "", "", "", "", "", "", "", "", "")
row.names(anova_results)[1] <- ""
anova_results[is.na(anova_results)] <- "--"
```

```{r echo=FALSE}
anova_results %>% knitr::kable("html",
               caption = 'Dimension 1 ANOVAs.') %>% 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = T) %>%
  kableExtra::add_header_above(c("", "DocuScope (Dim 1)" = 6, "Biber (Dim 1)" = 6))
```


```{r echo=FALSE}
fit1 <- lm(Factor2 ~ group, data = mda_micusp)
fit2 <- lm(Factor2 ~ group, data = mda_biber)

f1_aov <- data.frame(anova(fit1), r.squared = c(summary(fit1)$r.squared*100, NA))
f2_aov <- data.frame(anova(fit2), r.squared = c(summary(fit2)$r.squared*100, NA))

anova_results <- data.frame(rbind(c("DF", "Sum Sq", "Mean Sq", "F value", "Pr(>F)", "*R*^2^", 
                                    "DF", "Sum Sq", "Mean Sq", "F value", "Pr(>F)", "*R*^2^"), 
                                   cbind(round(f1_aov, 2), round(f2_aov, 2)))) 
colnames(anova_results) <- c("", "", "", "", "", "", "", "", "", "", "", "")
row.names(anova_results)[1] <- ""
anova_results[is.na(anova_results)] <- "--"

anova_results %>% knitr::kable("html",
               caption = 'Dimension 2 ANOVAs.') %>% 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = T) %>%
  kableExtra::add_header_above(c("", "DocuScope (Dim 2)" = 6, "Biber (Dim 2)" = 6))
```


```{r echo=FALSE}
fit1 <- lm(Factor3 ~ group, data = mda_micusp)
fit2 <- lm(Factor3 ~ group, data = mda_biber)

f1_aov <- data.frame(anova(fit1), r.squared = c(summary(fit1)$r.squared*100, NA))
f2_aov <- data.frame(anova(fit2), r.squared = c(summary(fit2)$r.squared*100, NA))

anova_results <- data.frame(rbind(c("DF", "Sum Sq", "Mean Sq", "F value", "Pr(>F)", "*R*^2^", 
                                    "DF", "Sum Sq", "Mean Sq", "F value", "Pr(>F)", "*R*^2^"), 
                                   cbind(round(f1_aov, 2), round(f2_aov, 2)))) 
colnames(anova_results) <- c("", "", "", "", "", "", "", "", "", "", "", "")
row.names(anova_results)[1] <- ""
anova_results[is.na(anova_results)] <- "--"

anova_results %>% knitr::kable("html",
               caption = 'Dimension 3 ANOVAs.') %>% 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = T) %>%
  kableExtra::add_header_above(c("", "DocuScope (Dim 3)" = 6, "Biber (Dim 3)" = 6))
```

## Place Disciplines along Dimensions

```{r}
p4 <- stickplot_dual(mda_micusp, mda_biber, n_factor = 1, title_a = "DocuScope", title_b = "Biber")
```

```{r fig.height=5, fig.width=7.5, echo=FALSE, fig.cap="Position of disciplines along Dimension 1."}
p4
```

```{r evaluate=FALSE, echo=FALSE}
ggsave("output/biber_1.png", p4,  height = 5, width = 7.5, units = "in", dpi = 300)
```


```{r}
p5 <- stickplot_dual(mda_micusp, mda_biber, n_factor = 2, title_a = "DocuScope", title_b = "Biber")
```

```{r fig.height=5, fig.width=7.5, echo=FALSE, fig.cap="Position of disciplines along Dimension 2."}
p5
```

```{r evaluate=FALSE, echo=FALSE}
ggsave("output/biber_2.png", p5,  height = 5, width = 7.5, units = "in", dpi = 300)
```

```{r}
p6 <- stickplot_dual(mda_micusp, mda_biber, n_factor = 3, title_a = "DocuScope", title_b = "Biber")
```

```{r fig.height=5, fig.width=7.5, echo=FALSE, fig.cap="Position of disciplines along Dimension 3."}
p6
```

```{r evaluate=FALSE, echo=FALSE}
ggsave("output/biber_3.png", p6,  height = 5, width = 7.5, units = "in", dpi = 300)
```



# Bibliography


